Towards a Neural Co-Processor Which Restores Movement After Stroke: Modeling a Proof-of-Concept

### 2021-11-11 paper/article.tex
Paper progress

The simulated circuit, from Michaels et al. \cite{michaels.mrnn}, was trained
to resemble the grasping circuits of monkey subjects engaged in a delayed reach-to-grasp task.
Its design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural circuits that have activation dynamics similar to natural circuits,
including delayed grasping tasks \cite{susillo.mrnn}. The Michaels circuit consists of a
``modular'' vanilla RNN (mRNN), and a linear readout layer. Each ``module'' consists of 100 vanilla
RNN neurons, with a non-linearity applied on the outputs. The modules are internally fully connected,
and are connected to each other sparsely (10\% connectivity). The inputs are visual features
intended to capture the view the monkeys had during the task, specifically VGGNet features from 3D
renderings of the same objects which the monkeys grasped. The outputs are muscle length
velocities for the should, arm, and hand of the monkey during the trial. The natural
velocities were captured with a motion capturing glove, and the artificial neural network
was trained to recapitulate those grasping motions. Data and trained models from this work
were supplied to us by the lead author Jonathan Michaels, for the purpose of our present
simulation. See Fig. \ref{fig:michaels}.

@@

\subsubsection{Stimulation model}
TODO: overview of spatio-temporal smoothing, connect to figure.

\subsubsection{Observation model}
TODO: overview of spatial / gaussian convolution concept, connect to $fig:stim_and_obs$.

\begin{figure}
\includegraphics[width=\textwidth]{stim_and_obs.png}
\caption{Example of stimulation and observation for a single trial}
\centering
\label{fig:stim_and_obs}
\end{figure}

\subsection{Experiments}
Summary of tunables:
\begin{itemize}
	\item Learning rate schedules
	\item Activation functions
	\item Obs function dimensionality
	\item Stim dimensionality
	\item Dimensionality of EN, CPN
\end{itemize}

Experiment types:
\begin{itemize}
	\item Connectivity lesion
	\item Co-adaptive with connectivity lesion
	\item Co-adaptive with partial M1 loss
	\item Co-adaptive with partial AIP loss
	\item Co-adaptive with data reuse
\end{itemize}

\subsection{Training the Co-Processor}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network.
The loss function when training the EN is the MSE loss between the actual and
predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN output, and target output, to the CPN. See Fig.
\ref{fig:training}.

@@

\begin{itemize}
	\item Train an EN
		\begin{itemize}
			\item Training the EN until predictive power reaches a threshold, based on current task performance
			\item Training regime based on most recent CPN, similar CPNS (noise added to params), and white noise stimulation
			\item Single batch of data from applying current CPN set and white noise to the brain, EN fit to ``offline'' it over maining epochs
			\item Predictive power alone is not sufficient for use in training the CPN: we need to form the batch as above, otherwise training is unstable.
			\item White noise and noised params
			\item ** Learning instability
			\item ** Predictive power isn't enough
			\item ** Illustrate this
		\end{itemize}
	\item Train the CPN
		\begin{itemize}
			\item Backprop through the EN, effectively using the EN's prediction as ground truth
		\end{itemize}

@@

\section{Results}
\begin{itemize}
	\item Task performance improves drastically.
		\begin{itemize}
			\item Fig: normalized task loss vs CPN training epoch
			\item Fig: normalized task loss vs training epoch
			\item Fig: normalized task loss vs training epoch
			\item Split by experiment
		\end{itemize}
	\item Fine tuning takes a long time
	\item Object classes separate. Fig: sep vs experiment vs epoch
	\item Can co-adapt with the brain, as it changes
	\item Training efficiency analysis. Show num epochs with and without recycling.
\end{itemize}

>>>>

The simulated circuit, from Michaels et al. \cite{michaels.mrnn}, was trained
to resemble the grasping circuits of monkey subjects engaged in a delayed reach-to-grasp task.
Its design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural circuits that have activation dynamics similar to natural circuits,
including delayed grasping tasks \cite{susillo.mrnn}. The Michaels circuit consists of a
``modular'' vanilla RNN (mRNN), and a linear readout layer. Each ``module'' consists of 100 vanilla
RNN neurons, with a non-linearity applied on the outputs. The modules are internally fully connected,
and are connected to each other sparsely (10\% connectivity). The inputs are visual features
intended to capture the view the monkeys had during the task, specifically VGGNet features from 3D
renderings of the same objects which the monkeys grasped. The outputs are muscle length
velocities for the should, arm, and hand of the monkey during the trial. The natural
velocities were captured with a motion capturing glove, and the artificial neural network
was trained to recapitulate those grasping motions. Data and trained models from this work
were supplied to us by the lead author Jonathan Michaels, for the purpose of our present
simulation. We re-implemented his models' logic in PyTorch, and loaded his trained parameters
into it, for one of his subjects, arbitrarily chosen. See Fig. \ref{fig:michaels}.

@@

\subsubsection{Stimulation model}

\begin{figure}[h]
	\includegraphics[width=0.45\textwidth]{stim_single.png}
	\caption{Spatial and temporal smoothing of a single 16 dimensional,
	         randomized $\theta$, onto 100 simulated neurons.
		 Color values indicate the magnitude of the value summed into
		 each neuron's hidden state in that time step. After $t=0$,
		 $\theta$ is the zero vector.}
	\centering
	\label{fig:stim_single}
\end{figure}

In order to inject another aspect of realism into our simulation, we subject
the co-processor to a stimulation model, rather than allowing it to directly
influence the hidden state of the simulated network's neurons. Our stimulation
model includes aspects of both spatial and temporal smoothing.

In our experiments, we stimulate only the output module, since this co-processor's
purpose is to improve external task performance, which it is able to do with
only stimulation of the output module of the network. It is conceivable that
a co-processor could stimulate other areas of the brain to improve task performance
downstream, or to probe the brain to better reveal the user's intent (i.e. object
shape), but we did not explore those possibilities in this paper.

The stimulation function $S$ receives as inputs the stimulation parameters $\theta$
provided by the CPN, which it accumulates in an internal memory. Each time step,
it outputs a modifier to the internal states of all neurons in the output module,
based on the current state of memory in the function object. That memory allows
the stimulation function to perform temporal smoothing. Specifically, we used a
simple exponential decay model where the function reads its current state of memory,
sums in the new parameters, and then decays each memory element towards $0.0$ at some
rate. Roughly speaking, this design intends to approximate the notion of dissipation: the
effect of stimulation is not instantaneous, but rather decays with time, as charge
dissipates into the surrounding area.

Likewise, the stimulation function maps the relatively low dimensional stimulation
parameter vector $\theta$ onto the M1 cells which it stimulates. In our experiments,
our stimulation parameters $\theta$ had 16 dimensions, which we found was
sufficient to allow significant improvement in task performance. The stimulation
function uses a simple Gaussian smoothing to map those parameters onto the simulated
M1 neurons. Each parameter represents, in a sense, an electrode located along a single
spatial dimension, whose stimulation affects the neurons in its vicinity more than
it affects others, according to each neuron's Mahalanobis distance from it. The neurons
are aligned along that dimension arbitrarily. We fix the width of the Gaussians
arbitrarily to $\sigma=1.75$. Note that we do not attempt to model the physical
layout of our simulated neurons, but simply want to disallow our co-processor
from having an unrealistically high degree of spatial resolution.

Thus, the governing equations of the stimulation become:

\begin{equation}
\alpha_{t} = \tau\alpha_{t-1} + \theta_{t}
\end{equation}
\begin{equation}
s_{t} = C(\alpha_{t})
\end{equation}

\begin{itemize}
	\item $\alpha$: the 16 dimensional internal activation, or memory
        of our stimulation
	\item $\tau$: our decay rate, which we set arbitrarily to $0.7$
	\item $C$: a precalculated $100 x 16$ matrix which provides our
	Gaussian smoothing
	\item $s_{t}$: the stimulation we apply to each neuron at the
	given time step.
\end{itemize}

The governing equations of the simulated network then become:

\begin{equation}
x_{t+1} = Jx_{t} + Iv_{t} + s_{t}
\end{equation}
\begin{equation}
a_{t} = tanh(x_{t})
\end{equation}
\begin{equation}
y_{t} = La_{t} + b
\end{equation}

\begin{itemize}
	\item $x$: the hidden state of each neuron
	\item $J$: the recurrence weight matrix
	\item $I$: the input response matrix
	\item $L$, $b$: the parameters of the linear readout
	\item $y$: the output of the network
\end{itemize}

See Fig. \ref{fig:stim_single} for a visual depiction of an example where
$\theta$ is non-zero at $t=0$, and zero for all other times, in order to see how
stimulation is applied across space and time. In Fig. \ref{fig:stim_and_obs},
we see stimulation applied during the simulation, in a real trial. In this case,
the trial is one where $50\%$ of M1 neurons are lesioned to have zero output.
Observations of the network's hidden state (explained in the next section) are taken
from the first two modules, and stimulation is applied to the last module. This fact
is intended to capture the difficulties with observing the same cortex that you are
stimulating, due to stimulation artifacts. The trial we show depicts stimulation
supplied by a highly trained co-processor, where task performance has been
largely restored. We explain this experiment in more detail below.

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{stim_and_obs.png}
	\caption{Example of stimulation and observation for a single trial. Here, M1
	has been lesioned 50\%, as seen in the zero (light blue) hidden states. We
	gather observations only from the AIP and F5 modules, as we explain in
	Section \ref{sec:experiments}. Stimulation is applied to M1, to drive the
	network output.}
	\label{fig:stim_and_obs}
\end{figure}

\subsubsection{Observation model}
Our observation model likewise relies on a notion of electrodes spread along
a single spatial dimension. As with stimulation, this is not intended to
capture true physical relationships between neurons and electrodes, but
rather to act as a dimensionality reduction that isn't designed to specifically
to favor task success, such as PCA. We treat the neurons of each module as
occupying their own space, with 20 electrodes arrayed along each.

We use a similar Gaussian-based approach as with stimulation, but in this
case our approach looks much like a Gaussian convolution, where the Gaussian
kernel is centered at each electrode position. Each `electrode' is read out
as a weighted average of all neurons in the given module, with weights being
provided by the Gaussian kernel. As above, refer to Fig. \ref{fig:stim_and_obs}
for a visual example.

\subsubsection{Brain co-adaptation}
To demonstrate the co-processor's ability to adapt to the non-stationarity
of the brain, we also change it throughout the co-processor's training.
Specifically, we simulate the brain's co-adaptation with the stimulation
to cooperatively solve the problem, as the stimulation is also changing
through the CPN's training.

We achieve this through a simple error backpropagation, using PyTorch's
implementation of the $Adam$ optimizer. With each trial, task loss is
calculated and backpropagated into the mRNN, allowing it to adapt to
its inputs, one of which is the stimulation being applied. The learning
rate for the optimizer is set arbitrarily to a relatively low rate of
$1e-7$.

\subsection{Training Algorithm}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network.
The loss function when training the EN is the MSE loss between the actual and
predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN output, and target output, to the CPN. See Fig.
\ref{fig:training}.

@@

Alternate training EN and CPN.
\begin{itemize}
	\item Train an EN
		\begin{itemize}
			\item Training the EN until predictive power reaches a threshold, based on current task performance
			\item Training regime based on most recent CPN, similar CPNS (noise added to params), and white noise stimulation
			\item Single batch of data from applying current CPN set and white noise to the brain, EN fit to ``offline'' it over maining epochs
			\item Predictive power alone is not sufficient for use in training the CPN: we need to form the batch as above, otherwise training is unstable.
			\item White noise and noised params
			\item ** Learning instability
			\item ** Predictive power isn't enough
			\item ** Illustrate this
		\end{itemize}
	\item Train the CPN
		\begin{itemize}
			\item Backprop through the EN, effectively using the EN's prediction as ground truth
		\end{itemize}

@@

\subsection{Experiments}
\label{sec:experiments}

Using the simulation described above, we perform three experiments to
demonstrate the co-processor's ability to learn:
\begin{itemize}
	\item 50\% AIP loss
	\item 100\% connectivity lesion between the F5 and M1 modules
	\item 50\% M1 loss
\end{itemize}

Additionally, we perform each of the ...

Summary of tunables:
\begin{itemize}
	\item Learning rate schedules
	\item Activation functions
	\item Obs function dimensionality
	\item Stim dimensionality
	\item Dimensionality of EN, CPN
\end{itemize}

\section{Results}

\begin{figure}[h]
\includegraphics[width=\textwidth]{results_standin.png}
\caption{Loss and class separation vs training epoch}
\centering
\label{fig:results}
\end{figure}

\begin{itemize}
	\item Task performance improves drastically.
		\begin{itemize}
			\item Fig: normalized task loss vs CPN training epoch
			\item Fig: normalized task loss vs training epoch
			\item Fig: normalized task loss vs training epoch
			\item Split by experiment
		\end{itemize}
	\item Fine tuning takes a long time
	\item Object classes separate. Fig: sep vs experiment vs epoch
	\item Can co-adapt with the brain, as it changes
	\item Training efficiency analysis. Show num epochs with and without recycling.
\end{itemize}
